Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis

### 2018-02-05 paper/finalish/article_1023.tex
update equation references to use eqref

To apply the general model of $f_{jh}$ to the campylobacteriosis data, assume $z$ is the categorical variable representing the classified location of each human case. Then two ways of treating the variable $z$ in model fitting are proposed: one is to treat it as numeric, and the other as categorical. To differentiate the performance between the two fitted models, we link ``the linear model'' and ``the categorical model'' to the first and the latter fitted model, respectively. Hence, the linear prediction function for source $j$ for each human case given the degree of rurality is a numeric variable and can be written as
\begin{align}
  f_{jh} = \alpha_{j} + \beta_{j} z_{h},
  \label{modL}
\end{align}
where $z_{h}$ can be any number of the seven scales if case $h$ was from such a degree of rurality. Conversely, if we treat each of the seven rurality degrees as an indicator with a superscript number $d$, which corresponds with the position of the category ranged from -3 to 3, the model~(\ref{modL}) can be rewritten as
\begin{equation}
  f_{jh} = \beta_{1j} z_{1h}+ \beta_{2j} z_{2h} + \ldots + \beta_{dj} z_{dh} + \ldots +  \beta_{7j} z_{7h},
  \label{modC}
\end{equation}
where
\begin{equation*}
z_{dh} =
\left\{\begin{array}{rcl}
1 && \mbox{if case $h$ is in the category $d$,} \notag \\ 0 && \mbox{otherwise}. 
\end{array}\right.
\end{equation*}
As a consequence, the estimated attribution probabilities are obtained via equation~(\ref{capF}) after fitting the data to model~\eqref{modL} or to model~\eqref{modC}.

\subsection{MCMC algorithm}
In the interest of quantifying the uncertainty of the posterior attribution probability, we perform Bayesian inference for source attribution probabilities using Markov chain Monte Carlo (MCMC) methods. Assume the priors on parameters of interest in model~(\ref{modL}) and model (4) follow a standard normal distribution. Let $\boldsymbol{\theta}$ denote the vector of parameters, with elements $\theta_{t}$ for $t=1, \ldots, T$. For example, $\boldsymbol{\theta}=\{\alpha_1, \alpha_2, \alpha_3, \beta_1, \beta_2, \beta_3\}$ in model~\eqref{modL} or $\boldsymbol{\theta}=\{\beta_{11}, \beta_{12}, \ldots, \beta_{dj}, \ldots, \beta_{73}\}$ in model~\eqref{modC}. We use the Metropolis-Hastings algorithm to update $\theta_{(t)}$ and hence $f_{jh}$ and $ F_{jh}$, in a Markov chain with a length of 30,000 iterations. The first 5,000 samples are removed as the burn-in period (during which time the chain converges) and the sequence is thinned every 10$^{th}$ sample to reduce computer storage. Here are the steps in detail:

@@

In light of the different estimates of the sampling distribution of genotypes from the asymmetric Island and the simple models, two approaches of model fitting (linear and non-linear) were utilized to estimate the attribution probability via the Metropolis-Hastings algorithm. Under the assumption of a logit scale on the attribution probability, we examined how the values of $f$ affected the resulting attribution probability. We changed the standard deviation of the prior distribution from 1 to 2 in the algorithm. The associated posterior mean of attribution probability for each source was invariant despite the fact that the proposals of $f$ jumped even wider in the random walk. This is because the denominator of the equation~(\ref{capF}) balances the impact made by the values of $f$. Further, we included a rurality covariate in the model fitting. Our findings not only coincided with the hypothesis that poultry is the major cause of human campylobacteriosis in urban areas \cite{MullA, Marsh, MullM, Leve} but also revealed the linear geographic effect on the infection. It means that poultry is less likely responsible for the infection when cases dwell away from urban areas. Furthermore, the result of decreasing poultry contribution in main urban areas from fitting the data with post-intervention period accorded with what Sears A. and other researchers investigated the effect of poultry intervention \cite{AnnS}. This could provide additional information that would help policy-makers to target interventions to populations residing in urban compared to rural areas.

>>>>

To apply the general model of $f_{jh}$ to the campylobacteriosis data, assume $z$ is the categorical variable representing the classified location of each human case. Then two ways of treating the variable $z$ in model fitting are proposed: one is to treat it as numeric, and the other as categorical. To differentiate the performance between the two fitted models, we link ``the linear model'' and ``the categorical model'' to the first and the latter fitted model, respectively. Hence, the linear prediction function for source $j$ for each human case given the degree of rurality is a numeric variable and can be written as
\begin{align}
  f_{jh} = \alpha_{j} + \beta_{j} z_{h},
  \label{modL}
\end{align}
where $z_{h}$ can be any number of the seven scales if case $h$ was from such a degree of rurality. Conversely, if we treat each of the seven rurality degrees as an indicator with a superscript number $d$, which corresponds with the position of the category ranged from -3 to 3, the model~\eqref{modL} can be rewritten as
\begin{equation}
  f_{jh} = \beta_{1j} z_{1h}+ \beta_{2j} z_{2h} + \ldots + \beta_{dj} z_{dh} + \ldots +  \beta_{7j} z_{7h},
  \label{modC}
\end{equation}
where
\begin{equation*}
z_{dh} =
\left\{\begin{array}{rcl}
1 && \mbox{if case $h$ is in the category $d$,} \notag \\ 0 && \mbox{otherwise}. 
\end{array}\right.
\end{equation*}
As a consequence, the estimated attribution probabilities are obtained via equation~\eqref{capF} after fitting the data to model~\eqref{modL} or to model~\eqref{modC}.

\subsection{MCMC algorithm}
In the interest of quantifying the uncertainty of the posterior attribution probability, we perform Bayesian inference for source attribution probabilities using Markov chain Monte Carlo (MCMC) methods. Assume the priors on parameters of interest in model~\eqref{modL} and model~\eqref{modC} follow a standard normal distribution. Let $\boldsymbol{\theta}$ denote the vector of parameters, with elements $\theta_{t}$ for $t=1, \ldots, T$. For example, $\boldsymbol{\theta}=\{\alpha_1, \alpha_2, \alpha_3, \beta_1, \beta_2, \beta_3\}$ in model~\eqref{modL} or $\boldsymbol{\theta}=\{\beta_{11}, \beta_{12}, \ldots, \beta_{dj}, \ldots, \beta_{73}\}$ in model~\eqref{modC}. We use the Metropolis-Hastings algorithm to update $\theta_{(t)}$ and hence $f_{jh}$ and $ F_{jh}$, in a Markov chain with a length of 30,000 iterations. The first 5,000 samples are removed as the burn-in period (during which time the chain converges) and the sequence is thinned every 10$^{th}$ sample to reduce computer storage. Here are the steps in detail:

@@

In light of the different estimates of the sampling distribution of genotypes from the asymmetric Island and the simple models, two approaches of model fitting (linear and non-linear) were utilized to estimate the attribution probability via the Metropolis-Hastings algorithm. Under the assumption of a logit scale on the attribution probability, we examined how the values of $f$ affected the resulting attribution probability. We changed the standard deviation of the prior distribution from 1 to 2 in the algorithm. The associated posterior mean of attribution probability for each source was invariant despite the fact that the proposals of $f$ jumped even wider in the random walk. This is because the denominator of the equation~\eqref{capF} balances the impact made by the values of $f$. Further, we included a rurality covariate in the model fitting. Our findings not only coincided with the hypothesis that poultry is the major cause of human campylobacteriosis in urban areas \cite{MullA, Marsh, MullM, Leve} but also revealed the linear geographic effect on the infection. It means that poultry is less likely responsible for the infection when cases dwell away from urban areas. Furthermore, the result of decreasing poultry contribution in main urban areas from fitting the data with post-intervention period accorded with what Sears A. and other researchers investigated the effect of poultry intervention \cite{AnnS}. This could provide additional information that would help policy-makers to target interventions to populations residing in urban compared to rural areas.
